A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-03 podc_paper/algorithm.tex
working on correctness of getEnqueue

As mentioned above, we are free to order operations within a block however we like.
For convenience, we order the enqueues and dequeues separately, and put the 
operations that were propagated from the left child before the operations from the right child.
More formally, we inductively define an sequence $E(B)$ of the enqueues represented by a block $B$.
If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is either the single
enqueue represented by the block (if $B$ stores an enqueue), or the empty sequence (if $B$ stores a dequeue).
If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s enqueue sequence is the concatenation $E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)$.
The dequeue sequence $D(B)$ of a block $B$ is defined symmetrically.  We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.

@@

Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
Thus, the direct subblocks of $v.\fld{blocks}[i]$ are
\begin{eqnarray}\label{defsubblock}
v.\fld{left.blocks}[v.\fld{blocks}[i-1].\fld{end\sub{left}}+1..v.\fld{blocks}[i].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
v.\fld{right.blocks}[v.\fld{blocks}[i-1].\fld{end\sub{right}}+1..v.\fld{blocks}[i].\fld{end\sub{right}}].
\end{eqnarray}
This allows us to navigate to the direct subblocks of any block easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $v.\fld{blocks}[i]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{i}]$.
These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue in the linearization ordering.
We use the \fld{sum\sub{enq}} fields of the root's blocks to do a binary search
to locate which block in the root contains the operation.
If we know which block $B$ in some \node\ $v$ contains the enqueue,
we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
and then to do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
the enqueue.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
All of the information needed for this search process can be derived from the 
\eleft, \eright\ and \fld{sum\sub{enq}} fields.

@@

    \If{\var{v} is a leaf node}
        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v}.\fld{left}}
            \State \fld{dir} \assign\ \fld{left}
        \Else
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf
        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t. $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} 
    \EndIf
\EndFunction{GetEnqueue}
\end{algorithmic}
\caption{\label{pseudocode2}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
\end{figure}

>>>>

As mentioned above, we are free to order operations within a block however we like.
For convenience, we order the enqueues and dequeues separately, and put the 
operations that were propagated from the left child before the operations from the right child.
More formally, we inductively define an sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
represented by a block $B$.
If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s operation sequences are defined by the concatenations 
\begin{eqnarray}
E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)\nonumber\\
D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
\end{eqnarray}
We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.

@@

Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
Thus, the direct subblocks of $v.\fld{blocks}[b]$ are
\begin{eqnarray}\label{defsubblock}
v.\fld{left.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{left}}+1..v.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
v.\fld{right.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{right}}+1..v.\fld{blocks}[b].\fld{end\sub{right}}].
\end{eqnarray}
This allows us to navigate to the direct subblocks of any block easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $v.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue in the linearization ordering.
We use the \fld{sum\sub{enq}} fields of the root's blocks to do a binary search
to locate which block in the root contains the operation.
If we know which block $B$ in some \node\ $v$ contains the enqueue,
we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
and then to do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
the enqueue.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
All of the information needed for this search process can be derived from the 
\eleft, \eright\ and \fld{sum\sub{enq}} fields.

@@

    \If{\var{v} is a leaf node}
        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v}.\fld{left}}
            \State \fld{dir} \assign\ \fld{left}
        \Else
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf \label{endChooseDir}
        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t. $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
    \EndIf
\EndFunction{GetEnqueue}
\end{algorithmic}
\caption{\label{pseudocode2}Queue implementation, continued.}
%\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}
\end{figure}

### 2023-01-03 podc_paper/correctness.tex
working on correctness of getEnqueue

\subsection{Correctness of GetEnqueue}
-------

\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.

\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
 From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order. Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing. By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node. Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$. Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.  Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$. Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}. By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$. Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
\end{proof}

\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
\end{lemma}
\begin{proof}
We will prove this lemma by induction on the node $n$'s height. For the base case, suppose $n$ is a leaf. Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation. Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.

For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.  From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}). By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$. So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}. By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
\begin{align*}
&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
 \end{align*}
First, we consider the case where the \nf{Enqueue} we are looking for is in the left child. There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

Otherwise, the \nf{Enqueue} we are looking for is in the right child. Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child. So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}). Other parameters for the right child are chosen similarly to the left child.

So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$. Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
\end{proof}

\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/blockSumEnq.png}

  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}. Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey. \nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
  }\label{figGet}
\end{figure}

>>>>

\subsection{Correctness of GetEnqueue}

\begin{lemma}\label{lem::get}
If $1\leq i\leq |E(v.\fld{blocks}[b])|$ then \op{getEnqueue}($v,b,i$) returns the argument of the $i$th enqueue in $E(v.\fld{blocks}[b])$.
\end{lemma}
\begin{proof}
We prove the claim by induction on the height of node $v$.
If $v$ is a leaf, the hypothesis implies that $i=1$ and the block $v.\fld{blocks}[b]$ represents 
an enqueue whose argument is stored in $v.\fld{blocks}[b].\fld{element}$.
\op{GetEnqueue} returns the argument of this enqueue at line \ref{getBaseCase}.

Assuming the claim holds for $v$'s children, we prove it for $v$.
Let $B$ be $v.\fld{blocks}[b]$.
By (\ref{defSeqs}),
$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
of $B$, which are listed in (\ref{defsubblock}).
By \Cref{lem::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
of enqueues in $E(B)$ that come from $B$'s subblocks in $v$'s left child.
Thus, $dir$ is set to the direction for the child of $v$ that contains the required enqueue.
Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
By \Cref{lem::sum}, $v.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
enqueues in $E'$ before the enqueues of block $B'$, so
the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
returns the required result, by the induction hypothesis.
\end{proof}

-------

### 2023-01-03 podc_paper/related.tex
working on correctness of getEnqueue

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.

Jayanti and Petrovic introduced a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
Our implementation uses their idea of having
a tournament tree among processes to agree on the linearization of
operations.

>>>>

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
We adapt their idea of processes using
a tournament tree to agree on a linearization of
operations.
